What We Eval in the Shadows

### 2021-08-12 paper/main.tex
Proofread 5.2

Table~\ref{tab:highlevelenvironments} shows that most calls evaluate in the
scope of a function, global is a distant second. This means that most reads and
writes act on local variables. But of which function? Table~\ref{tab:funoffset}
gives the offset of that function on the call stack: 0 is function that called
\eval, 1 is that function's caller, and so on. In 81\% of cases, \eval is
resolved the caller -- the variables of the function where the call to \eval
textually occurs are read and written. About 1.5\% of sites evaluate their
argument three frames up the call stack, or above. Modular reasoning is thus
impossible in R. Since the actions of \eval happen at a distance, understanding
any given code snippet requires knowing which functions may be called
transitively from that snippet. In \packageNbOneCategoryEnvirSitePercent of
sites, only kind of environment is observed.

@@

Each synthetic environment has parent that is specified when calling
\k{new.env}. When \k{envir} is a list or a data frame, \eval uses its third
argument (\k{enclos}). The parent is used to search for variables not found in
its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
kinds for synthetic environments. Most of them are functions.

Global \evals split between intentional and accidental one. Direct references to
the top-level, using \k{globalenv()} or \k{.GlobalEnv}, are rare; they occur in
only \packageNbExplicitGlobalSites sites. Thus we suspect most uses of global
are accidental. They arise from the fact that our corpus, consists of many code
snippet that are run at top-level, global is thus the caller of \eval. This
noteworthy because writes to global variables are visible to all functions and
are not reclaimed by the garbage collector. So accidental uses may pollute that
namespace. Table~\ref{tab:polyenvir} gives the number of kinds seen at a given
call site.

\paragraph{Discussion.}
The data presented here is not surprising. The main use of \eval is to provide a
customizable extension mechanism for the behavior of functions. They are a way
to parameterize the function with any behavior that can be expressed in R. The
behavior of \eval within its enclosing function is to read and write variables,
mostly read, and very rarely delete variables. There are also some cases where
new variables are injected in a function. Usually, this in the direct caller,
but can sometimes take effect several frames up the call stack. There is
something brittle about code relying on the position of a caller, many
refactorings may break code that does that.

One bit of information that we lack is how the environment was obtained. Our
expectation is that the expression was extracted from a promise using
\k{substitute}, perhaps modified, and then evaluated with the environment coming
from the same promise. We believe the case where \eval is provided the results
of programmatically selecting some call stack to be less frequent.

Synthetic environments are relatively frequent. Common uses-case include
evaluation of an expression in a sandbox, or using a data structure as
environment, for example one could evaluate a method of an object and use and
environment to hold the object's fields.

We have already explained the popularity of global environments, as for package
environments, they are used in only \packageNbPackageNamespaceEnvSites sites.
This is probably for the best as mutating the bindings of a loaded package is
frowned upon and R tries to discourage it.

>>>>

Table~\ref{tab:highlevelenvironments} shows that most calls evaluate in the
scope of a function, global is a distant second. This means that most reads and
writes act on local variables. But of which function? Table~\ref{tab:funoffset}
gives the offset of that function on the call stack: 0 is the function that called
\eval, 1 is that function's caller, and so on. In 81\% of cases, \eval uses
its caller's environment -- the variables of the function where the call to \eval
textually occurs are read and written. About 1.5\% of sites evaluate their
argument three frames up the call stack, or above. Modular reasoning is thus
impossible in R. Since the actions of \eval happen at a distance, understanding
any given code snippet requires knowing which functions may be called
transitively from that snippet. In \packageNbOneCategoryEnvirSitePercent of
the sites, only one kind of environment is observed.

@@

Each synthetic environment has a parent that is specified when calling
\k{new.env}. When \k{envir} is a list or a data frame, \eval uses its third
argument (\k{enclos}). The parent is used to search for variables not found in
its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
kinds for synthetic environments. Most of them are functions.

Global \evals split between intentional and accidental one. Direct references to
the top-level, using \k{globalenv()} or \k{.GlobalEnv}, are rare; they occur in
only \packageNbExplicitGlobalSites sites. Thus we suspect most uses of global
are accidental. They arise from the fact that our corpus consists of many code
snippets that are run at the top-level, global is thus the caller of \eval. This is 
noteworthy because writes to global variables are visible to all functions and
are not reclaimed by the garbage collector. So accidental uses may pollute that
namespace. Table~\ref{tab:polyenvir} gives the number of kinds seen at a given
call site.

\paragraph{Discussion.}
The data presented here is not surprising. The main use of \eval is to provide a
customizable extension mechanism for the behavior of functions. They are a way
to parameterize the function with any behavior that can be expressed in R. The
behavior of \eval within its enclosing function is to read and write variables,
mostly read, and very rarely, delete variables. There are also some cases where
new variables are injected in a function. Usually, this in the direct caller,
but it can sometimes take effect several frames up the call stack. There is
something brittle about code relying on the position of a caller, a lot of
refactoring may break code that does that.

One bit of information that we lack is how the environment was obtained. Our
expectation is that the expression was extracted from a promise using
\k{substitute}, perhaps modified, and then evaluated with the environment coming
from the same promise. We believe the case where \eval is provided with the results
of programmatically selecting some call stack to be less frequent.

Synthetic environments are relatively frequent. Common uses-case include
evaluation of an expression in a sandbox, or using a data structure as
environment. For example one could evaluate a method of an object and use an
environment to hold the object's fields.

We have already explained the popularity of global environments. As for package
environments, they are used in only \packageNbPackageNamespaceEnvSites sites.
This is probably for the best as mutating the bindings of a loaded package is
frowned upon and R tries to discourage it.
